---
title: "Practical machine Learning Project"
author: "William Gourley"
date: "December 9, 2015"
output: html_document
---

## Executive Summary
Two machine learning models were built in order to predict 20 test cases. The random forest model achieved an accuracy (99.77%), while the boosted model achieved an accuracy of (96.78%). Both models predicted the test cases identically therefore the boosted model was selected to be the final model as the model requires much less computation time and would therefore be more efficient in a Production environment.

The code illustrating the analysis is shown below.

## Overview
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways

## Load required Packages
```{r,LoadPackages,message=FALSE,warning=FALSE}
library(caret)
```

## Loading the Data
```{r,LoadData,message=FALSE,warning=FALSE}
#if data does not exist, then get it from the url's
if(!file.exists("./training.csv")) {
    #get the training data
    url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
    training <- read.csv(url(url))

    #write the training data
    write.csv(training, file = "./training.csv")

    #get the test data
    url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
    testing <- read.csv(url(url))

    #write the test data
    write.csv(testing,file="./testing.csv")
}

#read in training and test data from working directory
training <- read.csv("./training.csv",header=TRUE)
testing <- read.csv("./testing.csv",header=TRUE)

dim(training)
dim(testing)
```

## Data PreProcessing
The following preprocessing steps were carried out to clean up the data and reduce the number of predictor variables.

* All variables which had NA's were removed
* All variables which had near-zero variances were removed
* The first 7 variables were removed as they were purely descriptive and had no predictive value

The training data was thus reduced from 161 variables to 53 variables, 1 outcome variable (classe) and 52 predictor variables.

```{r,PreProcessData,message=FALSE,warning=FALSE}
#clean up training dataset
#remove columns with NA#s and near-zero variances
training.clean <- training[,sapply(training,function(x)!any(is.na(x)))]
#examine near zero variances
nzv <- nearZeroVar(training.clean,saveMetrics = TRUE)
subset(nzv,nzv == TRUE)
training.clean <- training.clean[,nzv$nzv==FALSE]

#remove columns 1:7 as they have no predictive value
training.clean <- training.clean[,-c(1:7)]
dim(training.clean)
```

## Create Training and Validation Datasets
The training set was split up into a train set which contained 60% of the observations, and a validation set containing 40% of the observations. The testing set was only used once at the end to predict the 20 test cases.

```{r,CreateTraining,message=FALSE,warning=FALSE}
#split training.clean into train (60%) and validate (40%) data sets
inTrain <- createDataPartition(y = training.clean$classe,p=0.60,list=FALSE)
train <- training.clean[inTrain,]
validate <- training.clean[-inTrain,]
dim(train)
dim(validate)
```

## Build Models
Two models were built, a random forest model and a boosted model. Both models were built using 10 'repeatedcv' cross-validations repeated 3 times - a total of 30 cross-validations.


The out-of-sample errors were estimated using the validation set and calculated using repeated cross-validations for both models.


The out-of-sample error for the Random Forest and Boosted models are 0.33% and 3.29% respectively.

```{r,BuildModels}
set.seed(12345)
```

### Random Forest
```{r,RandomForest,cache=TRUE,message=FALSE,warning=FALSE}
#build random forest model
if(!file.exists("./model.rf.RData")) {
    control <- trainControl(method = "repeatedcv",number = 10,repeats = 3,allowParallel = TRUE)
    model.rf <- train(classe ~ .,data=train,method="rf",prox=TRUE,trControl=control)
    #save model
    save(model.rf,file="./model.rf.RData")
} else {
    load(file="./model.rf.RData")
}
pred.rf <- predict(model.rf,validate)
validate$predCorrect <- pred.rf == validate$classe

confusionMatrix(validate$classe,pred.rf)

```

### Boosted Model
```{r,Boosted,cache=TRUE,message=FALSE,warning=FALSE}
#build boosted model
if(!file.exists("./model.gbm.RData")) {
    control <- trainControl(method = "repeatedcv",number = 10,repeats = 3,allowParallel = TRUE)
    model.gbm <- train(classe ~ .,data=train,method="gbm",trControl = control,verbose=FALSE)
    save(model.gbm,file="./model.gbm.RData")
} else {
    load(file="./model.gbm.RData")
}

pred.gbm <- predict(model.gbm,validate)
validate$predCorrect <- pred.gbm == validate$classe

confusionMatrix(validate$classe,pred.gbm)
```

## Test Cases Prediction
```{r,TestCasePrediction,message=FALSE,warning=FALSE}
pml_write_files <- function(x){
    n = length(x)
    for(i in 1:n){
        filename = paste0("problem_id_",i,".txt")
        write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
    }
}

pred.submission.rf <- predict(model.rf,testing)
answers.rf <- as.character(pred.submission.rf)
print(answers.rf)

pred.submission.gbm <- predict(model.gbm,testing)
answers.gbm <- as.character(pred.submission.gbm)
print(answers.gbm)

```

## Final Model Selection
The test case prediction results for both models are identical, therefore the boosted model was chosen as the final model, sacrificing a little accuracy for reduced computation build time.

```{r,ModelSelection,message=FALSE,warning=FALSE}
#use boosted model as final model as the computation is less
final.model <- model.gbm

#write submission files
setwd("./pml_submission")
pml_write_files(answers.gbm)
```